Implementing Distributed Systems Directly\\with I/O Automata in ioa++

### 2011-09-27 papers/paper1/related_work.tex
General revision and added conclusion.

This work grew out of the frustration encountered when trying to develop distributed systems with threads.
Lee elaborates on the problems faced when using threads for concurrency in~\cite{lee2006problem}.
These frustrations will increase as concurrent programming is the only foreseeable way to increase performance given the leveling trend of clock frequencies and the advent of multi-core processors\cite{sutter2005software}.

The idea to implement a formal model to simplify concurrent programming was inspired by the application of structured programming~\cite{dijkstra1968letters} to programming language design.
The I/O automata model of Lynch and Tuttle\cite{lynch1987hierarchical} was selected because it supports features we believe to be fundamental to developing distributed systems.
\cite{lynch1996distributed} shows how I/O automata can be applied to a number of problems in distributed computing.
Event-based proof techniques~\cite{manna1992temporal},~\cite{lamport1978time} can be used to reason about I/O automata.

Comparing I/O automata to every model for asynchronous, concurrent, and distributed computing is beyond the scope of this paper so we focus on exemplars.
The I/O automata model is based on state transitions which matches the imperative style that dominates modern concurrent programming.
This is similar to the motivation for UNITY~\cite{chandy1988parallel} and is in contrast to functional models such as the $\pi$-calculus~\cite{milner1992calculus} and actors~\cite{agha1986actors}.
The I/O automata model assumes independent state with explicit communication like communicating sequential processes (CSP)~\cite{hoare1978communicating} but unlike threads~\cite{lee2006problem} and the UNITY model~\cite{chandy1988parallel}.
All models for asynchronous, concurrent, and distributed computing admit non-determinism.
Lee points out that the difficulties of thread-based programming come from a need to ``prune'' non-determinism when paired with shared state and adopts the view that software components should be deterministic with respect to concurrency save a few that introduce non-determinism in a controlled way~\cite{lee2006problem}.
Programs in UNITY are composed by concatenating program texts and resolving shared variables~\cite{chandy1988parallel}.
For this reason, they share many of the same problems as threads under composition.
I/O automata can be viewed as taking the alternate approach by admitting non-determinism while prohibiting shared state.
In I/O automata, the sequential flow control of threads~\cite{lee2006problem} and CSP~\cite{hoare1978communicating} is replaced by the non-deterministic execution of conditional atomic actions which resembles execution in the UNITY model~\cite{chandy1988parallel}.
The ability to create new automata can be compared to actor creation in the actor model~\cite{agha1986actors}.
Actors communicate asynchronously by name using a buffered mail system while I/O automata communicate synchronously and anonymously according to composition.

To our knowledge, only the Spectrum Simulation System of Goldman~\cite{goldman1990distributed} and the IOA language developed by Lynch et al.~\cite{garland2003ioa} are the only existing systems that allow one to execute a system expressed as a collection of I/O automata.
Spectrum focuses on algorithm development through simulation.
IOA focuses on formal software development including both simulation and verification.

ioa++ can be compared directly to the IOA language due to the addition of a compiler~\cite{tsai2002code},~\cite{tauber2004verifiable},~\cite{tauber2004compiling} that has been used to implement a number of protocols~\cite{georgiou2009automated}.
\cite{georgiou2009automated} lists six challenges when developing the compiler: ``Program structuring,'' ``IOA Programs and external services,'' ``Modeling procedure calls,'' ``Composing automata,'' ``Nondeterminism'', and ``Implementing datatypes.''
Programming written in IOA are compiled to Java and executed on hosts communicating using the MPI library.
IOA requires programs to presented in a ``node-channel'' form that allows them to be composed with built in ``mediators'' that implement external services such as MPI.
IOA lacks support for procedure calls which complicates modeling and interfacing with external services; especially when a procedure call blocks.
The IOA compiler composes automata statically using a ``composer'' resulting in a ``node automaton.''
The next action to be executed is determined by a schedule function that resolves the non-determinism in the schedule.

IOA and ioa++ represent two fundamentally different approaches to programming directly with I/O automata.
The development of a new language was necessary for IOA due to their focus on formal verification.
ioa++ on the other hand uses the existing C++ language and compilers.
IOA encapsulates external services using built in ``mediators'' while ioa++ exposes operating system services via file descriptors relying on the programmer to ensure non-blocking operation.
IOA focuses on static composition to produce a single node program while ioa++ focuses on dynamic composition and allows concurrent execution.
Both IOA and ioa++ require the programmer to add code to schedule actions.

>>>>

This work grew out of the challenges faced when trying to develop distributed systems with threads.
Lee elaborates on the problems faced when using threads for concurrency in~\cite{lee2006problem}.
These same concerns are relevant to trends towards increasingly concurrent programming to increase performance, given the leveling trend of clock frequencies and the advent of multi-core processors\cite{sutter2005software}.

The direction implementation of a formal model to simplify concurrent programming was inspired by the application of structured programming~\cite{dijkstra1968letters} to programming language design.
The I/O automata model of Lynch and Tuttle\cite{lynch1987hierarchical} was selected because it supports features essential to developing distributed systems~\cite{lynch1996distributed}.
Temporal logic proof techniques~\cite{manna1992temporal},~\cite{lamport1978time} also can be used to reason about I/O automata, towards developing \emph{correct} systems.

Comparing I/O automata to every model for asynchronous, concurrent, and distributed computing is beyond the scope of this paper so we focus on exemplars.
The I/O automata model is based on state transitions which matches the imperative style that dominates modern concurrent programming.
This is similar to the motivation for UNITY~\cite{chandy1988parallel} and is in contrast to functional models such as the $\pi$-calculus~\cite{milner1992calculus} and actors~\cite{agha1986actors}.
The I/O automata model assumes independent state with explicit communication like communicating sequential processes (CSP)~\cite{hoare1978communicating} but unlike threads~\cite{lee2006problem} and the UNITY model~\cite{chandy1988parallel}.
All models for asynchronous, concurrent, and distributed computing admit non-determinism.
Lee points out that the difficulties of thread-based programming come from a need to ``prune'' non-determinism when paired with shared state and adopts the view that software components should be deterministic with respect to concurrency save a few that introduce non-determinism in a controlled way~\cite{lee2006problem}.
Programs in UNITY are composed by concatenating program texts and resolving shared variables~\cite{chandy1988parallel}.
For this reason, they share many of the same problems as threads under composition.
I/O automata can be viewed as taking the alternate approach by admitting non-determinism while prohibiting shared state.
In I/O automata, the sequential flow control seen with threads~\cite{lee2006problem} and CSP~\cite{hoare1978communicating} is replaced by the non-deterministic execution of conditional atomic actions which resembles execution in the UNITY model~\cite{chandy1988parallel}.
The ability to create new automata can be compared to actor creation in the actor model~\cite{agha1986actors}.
Actors communicate by name using a buffered mail system while communications in I/O automata are not buffered and anonymous.

To our knowledge, the Spectrum Simulation System~\cite{goldman1990distributed} and the IOA language~\cite{garland2003ioa} are the only existing systems that allow one to execute a system expressed as a collection of I/O automata.
Spectrum focuses on algorithm development through simulation.
IOA focuses on formal software development including both simulation and verification.

The ioa++ framework can be compared directly to the IOA language due to the addition of a compiler~\cite{tsai2002code},~\cite{tauber2004verifiable},~\cite{tauber2004compiling} that has been used to implement a number of protocols~\cite{georgiou2009automated}.
\cite{georgiou2009automated} lists six challenges when developing the compiler: ``Program structuring,'' ``IOA Programs and external services,'' ``Modeling procedure calls,'' ``Composing automata,'' ``Nondeterminism'', and ``Implementing datatypes.''
Programming written in IOA are compiled to Java and executed on hosts communicating using the MPI library.
IOA requires programs to presented in a ``node-channel'' form that allows them to be composed with built in ``mediators'' that implement external services such as MPI.
IOA lacks support for procedure calls which complicates modeling and interfacing with external services; especially when a procedure call blocks.
The IOA compiler composes automata statically using a ``composer'' resulting in a ``node automaton.''
The next action to be executed is determined by a schedule function that resolves the non-determinism in the schedule.

IOA and ioa++ represent two fundamentally different approaches to programming directly with I/O automata.
The development of a new language was necessary for IOA due to their focus on formal verification.
ioa++ on the other hand uses the existing C++ language and compilers.
IOA encapsulates external services using built in ``mediators'' while ioa++ exposes operating system services via file descriptors relying on the programmer to ensure non-blocking operation.
IOA focuses on static composition to produce a single node program while ioa++ focuses on dynamic composition and allows concurrent execution.
Both IOA and ioa++ require the programmer to add code to schedule actions and find this to be a common source of errors.
